---
title: "Probability Distributions"
subtitle: "MA336 Statistics<br/><br/>"
author: "Fei Ye <br/><br/> Department of Mathematics and Computer Science<br/><br/>"
date: '`r format(Sys.time(), "%B, %Y")`'
output:
  xaringan::moon_reader:
    css: [default, default-fonts, "Assets/css/custom.css"]
    mathjax: "https://cdnjs.cloudflare.com/ajax/libs/mathjax/2.7.4/MathJax.js?config=TeX-MML-AM_HTMLorMML-full"
    nature:
      ratio: '16:9' # alternatives '16:9' or '4:3' or others e.g. 13:9
      highlightLines: true
      highlightStyle: github
      countIncrementalSlides: false  
      beforeInit: ["Assets/js/resize.js"]
---

```{r, echo=FALSE}
knitr::opts_chunk$set(comment="#", fig.retina=2, crayon.enabled = TRUE, cache=TRUE)
set.seed(15)
library(stats)
library(kableExtra)
library(knitr)
library(formattable)
library(ggplot2)
library(ggthemes)
library(ggExtra)
library(dplyr,warn.conflicts = FALSE)
library(animation)
library(gifski)
library(latex2exp)
```

# Textbooks

- [Concepts in Statistics](https://courses.lumenlearning.com/wmopen-concepts-statistics/)

- [Introductory Statistics](https://open.umn.edu/opentextbooks/textbooks/introductory-statistics)

---
class: center middle

# Learning Goals

---

## Learning Goals for Probability and Probability Distribution

- Demonstrate understanding of random variables (discrete or continuous)

- Demonstrate understanding of characteristics of binomial and normal distributions.

- Calculate accurate probabilities of discrete or continuous random variables and interpret them in a variety of settings: normal distribution and binomial distribution.

- Calculate the standardized value (or $z$-score).

---

## Random Variables

- A **random variable**, usually written $X$, is a variable whose values are numerical quantities of possible outcomes a random experiment.

- A **discrete random variable** takes on only a finite or countable number of distinct values.

  **Example:** Roll a fair dice, the number of dots on the top faces is a discrete random variables takes on the possible values: 1, 2, 3, ,4, 5, 6.

- A **continuous random variable** takes on values which form an interval of numbers.

  **Example:** The height of an adult which takes on values in a certain interval.

---

## Probability Distributions (1/2)

- The **probability distribution** of a discrete random variable $X$ can be characterized a list of all possible values $x$ together with probabilities $P(x)$.

- A random variable $X$ (discrete and continuous) always has a **cumulative distribution function**: $F_X(x)=P(X\leq x)$ (= $\sum_{x_i\leq x} P(x_i)$ if $X$ is discrete).

- Recall basic rules of probability:
  - $0\leq P(X=x)\leq 1$.
  - the sum of all the probabilities is 1, that is $P(X\leq x_{max})=1$.
  - In particular, $0\leq F_X(x)\leq 1$.

- The probability distribution can be recovered from its cumulative distribution function.

---

## Probability Distributions (2/2)

- **Example:** Let $X$ be the number of heads that are observed when tossing two fair coins. The probability distribution can be characterized by the following table:
.center[
| $X$   | 0   | 1   | 2   |
|-----|-----|-----|-----|
| $P(X)$ |0.25 | 0.5 | 0.25|
]

  From the table, we may find the following cumulative distributions:
  $$P(X\leq 1)=P(X=0)+P(X=1)=0.25+0.5=0.75.$$
  $$P(X\leq 2)= P(X=0)+P(X=1)+P(X=2)=0.25+0.5+0.25=1.$$
  We may also recover probability distributions from the cumulative distribution, e.g.
  $$P(X=2)=P(X\leq 2)-P(X\leq 1).$$

---

## Mean and Standard Deviation of a Discrete Random Variable (1/3)

- The **mean** $\mu$ (also called the **expected value** $E(X)$) of a discrete random variable $X$ is the number $$\mu= E(X)=\sum xP(x).$$

- The **variance** $\sigma^2$ of a discrete random variable $X$ is the number $$\sigma^2=\sum (x-\mu)^2P(X).$$

- The **standard deviation** $\sigma$ of a discrete random variable $X$ is the square root of its variance: $$\sigma=\sqrt{\sum (x-\mu)^2P(X)}.$$

---

## Mean and Standard Deviation of a Discrete Random Variable (2/3)

- **Example:** One thousand raffle tickets are sold for $2 each. Each has an equal chance of winning. First prize is $500, second prize is $300, and third prize is $100. Find the expected value of gain, and interpret its meaning.

  **Solution:** Let $X$ denote the net gain from purchasing one ticket. The probability distribution for $X$ is
.center[
| $X$    | 498    | 298    | 98     | -2       |
|------|--------|--------|--------|----------|
| $P(X)$ | $\frac{1}{1000}$ | $\frac{1}{1000}$ | $\frac{1}{1000}$ | $\frac{997}{1000}$ |
]

  The expected gain is $$E(X)= 498\cdot \frac{1}{1000}+ 298\cdot\frac{1}{1000}+98\cdot \frac{1}{1000}+(-2)\cdot \frac{997}{1000}=-1.1,$$ which means that when buying one ticket, the buyer may expect a loss of $1.1.
  
---

## Mean and Standard Deviation of a Discrete Random Variable (3/3)

- **Example:** The wait times (rounded to multiples of 5) in the cafeteria at a Community College has the following probability distribution. Find the standard deviation.
.center[
| $X$ (minutes) | 5        | 10       | 15      | 20       | 25       |
| -------------------- | -------- | -------- | -------- | -------- | -------- |
| $P(X)$             | 0.13 | 0.25 | 0.31 | 0.21 | 0.1 |
]

  **Solution:** The mean of the distribution is $$ \mu= 5\cdot 0.13 +10\cdot 0.25+15\cdot 0.31+20\cdot 0.21 + 25\cdot 0.1 = 14.5.$$ The standard deviation is then $$\scriptstyle \sigma=\sqrt{(5-14.5)^2\cdot 0.13 +(10-14.5)^2\cdot 0.25+(15-14.5)^2\cdot 0.31+(20-14.5)^2\cdot 0.21 + (25-14.5)^2\cdot 0.1}\approx 5.9.$$

---

## Binomial Distribution (1/5)

- A **binomial experiment** is a probability experiment satisfying:

  1. The experiment has a fixed number $n$ of independent trials.
  2. Each trial has only two possible outcomes: a success (S) or a failure (F).
  3. The probability $p$ of a success is the same for each trial.

- The discrete random variable $X$ counting the number of successes in the $n$ trials is the **binomial random variable**. We say $X$ has a **binomial distribution** with parameters $n$ and $p$ and write it as $X\sim B(n, p)$.

- For $X\sim B(n, p)$, the **probability of getting exactly $x$ successes in $n$ trials** is $$P(X=x)=B(x,n,p)={_n C_x} p^x(1-p)^{n-x}=\frac{n!}{(n-x)!x!}p^x(1-p)^{n-x}.$$

- The notation $n!=n(n-1)\cdots 1$ is read as $n$ factorial. We set $0!=1.$

- The notation ${_n C_x}=\frac{n!}{(n-x)!x!}$ is read as $n$ choose $x$, which is the number of ways to choose $x$ objects from a set of $n$ objects.

---

## Binomial Distribution (2/5)

- **Example:** A card is selected from a standard deck and replaced. This experiment is repeated a total of $5$ times.
  - Find the probability of selecting exactly $3$ clubs.
  - Find the probability of getting at least $3$ clubs.
  
  **Solution:** This is a binomial experiment. The number to total trial is $n=5$. The number of success is $3$. The chance of a success is $p=\frac{13}{52}=\frac14$. Apply the binomial probability formula, we have
  $$P(X=3)=\frac{5!}{3!2!} \left(\frac{1}{4}\right)^3\left(\frac34\right)^2=10\cdot\frac{9}{4^5}\approx 0.088.$$

  To probability of getting at least $3$ club is
  $$P(X\geq 3) =1-P(X\leq 2)=1-(P(0)+P(1)+P(2))\approx 1-0.8965=0.1035.$$

---

## Binomial Distribution (3/5)
  
  **Solution:(continued)**
  To calculate $P(X\leq 2)$, we may also use the binomial distribution table or the Excel function `BINOM.DIST()`.

  **Method 1:** As $n=5$ and $p=0.25$, we use the following portion of the cumulative binomial distribution table.
  .center[Binomial Probability Table *n=5*

| n    | x    | 0.1    | 0.15   | 0.2    | 0.25   | 0.3    | 0.35   | 0.4    |
| ---- | ---- | ------ | ------ | ------ | ------ | ------ | ------ | ------ |
| 5    | 0    | 0.5905 | 0.4437 | 0.3277 | .red[0.2373] | 0.1681 | 0.116  | 0.0778 |
| 5    | 1    | 0.3281 | 0.3915 | 0.4096 | .red[0.3955] | 0.3602 | 0.3124 | 0.2592 |
| 5    | 2    | 0.0729 | 0.1382 | 0.2048 | .red[0.2637] | 0.3087 | 0.3364 | 0.3456 |
| 5    | 3    | 0.0081 | 0.0244 | 0.0512 | 0.0879 | 0.1323 | 0.1811 | 0.2304 |
  ]
  
  $P(X\le 2) \approx 0.2373+0.3955+0.2637= 0.8965.$

  **Method 2:** In Excel, $P(X\le 2)$ `=BINOM.DIST(2,5,0.25,TRUE)` $\approx 0.8965$.

???
In a calculator, use `binompdf(n, p, x)` for $P(X=x)$ and `binomcdf(n, p, x)` for $P(X\leq x)$.

---

## Binomial Distribution (4/5)

- The mean of a binomial distribution of $n$ trials is $$\mu =\sum xP(x)=\sum x\cdot \dfrac{n!}{(n-x)!x!}p^x(1-p)^{n-x} = np.$$

- The variance of a binomial distribution of $n$ trials is $$\sigma^2 =\sum (x-np)^2P(x)=\sum x^2P(x)-(np)^2=np(1-p).$$  

- The variance of a binomial distribution of $n$ trials is $$\sigma=\sqrt{np(1-p)}.$$

- We consider an event $E$ **unusual** if the probability $P(E)\leq 5\%$.

---

## Binomial Distribution (5/5)

- **Example:** The probability that an egg in a retail package is cracked or broken is 0.02.
  - Find the average number of cracked or broken eggs in a one dozen carton.
  - Find the standard deviation. Is getting at least two broken eggs unusual?

  **Solution:** Since there are 12 eggs and the chance of getting a cracked egg is 0.02, the average number of cracked is
  $$\mu =np=12\cdot 0.02=0.24.$$
  
  The standard deviation is
  $$\sigma=\sqrt{12\cdot 0.02\cdot(1-0.02)}\approx 0.4850.$$

  Recall the Empirical rule: 95% data are within 2 standard deviation away from the mean. Since $2>0.24+2\cdot 0.4850$, the chance of getting at least two cracked eggs is less than 5%, which is considered as unusual.

---

## Continuous Probability Distribution

- The probability distribution of a continuous random variable $X$ is characterized by its **probability density function** $f(X)$ satisfying that the probability $P(a\leq X\leq b)$ equals the area above the interval $[a, b]$ but under the graph of the function $f(X)$.

.center[

```{r echo=FALSE, fig.width=5, fig.asp=0.5}
puniform <- ggplot(
    data = data.frame(x = c(-0.5, 0.5)), aes(x)
  ) +
  stat_function(
    fun = dunif,
    args = list(min = -0.5, max = 0.5),
    colour = "blue"
  ) +
  xlab("X") +
  ylab("") +
  scale_y_continuous(breaks=NULL) +
  theme_bw() +
  scale_x_continuous(
    breaks = c(-0.35, 0, 0.35), 
    labels =c( "a", "mean", "b" )
  ) +
  stat_function(
    fun=dunif,
    args=list(min = -0.5, max = 0.5),
    xlim=c(-0.35, 0.35),
    geom="area",
    fill="#84CA72",
    alpha=0.2
  ) +
  ggtitle("A Uniform Distribution") +
  geom_hline(yintercept=0)+ 
  geom_segment(
    aes(
      x = 0,
      y = 0,
      xend = 0,
      yend = 1,
    ),
    color="#389912",
    linetype="dashed"
  ) +
  geom_segment(
    aes(
      x = -0.5,
      y = 0,
      xend = -0.5,
      yend = 1,
    ),
    linetype="dashed"
  ) +
  geom_segment(
    aes(
      x = 0.5,
      y = 0,
      xend = 0.5,
      yend = 1,
    ),
    linetype="dashed"
  ) +
  theme(
    panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
    plot.title = element_text(hjust = 0.5, size=24), panel.border = element_blank(),
    axis.title.x = element_text(
      margin = margin(t = -20),
      hjust=0.95,
      size=24
    ),
    axis.text.x = element_text(size=28)
  ) +
  annotate(
    geom='text',
    x=0,
    y=0.5,
    label=TeX("$P(a\\leq X\\leq b)$",
              output='character'),
    parse=TRUE,
    size = 8
  )

puniform
```
```{r echo=FALSE, fig.width=5, fig.asp=0.5}
pnormal <- ggplot(
    data = data.frame(x = c(-3, 3)), aes(x)
  ) +
  stat_function(
    fun = dnorm,
    args = list(mean = 0, sd = 1),
    colour = "blue"
  ) +
  xlab("X") +
  ylab("") +
  scale_y_continuous(breaks=NULL) +
  theme_bw() +
  scale_x_continuous(
    breaks = c(-1.5, 0, 1.2),
    labels =c( "a", "mean", "b" )
  ) +
  stat_function(
    fun=dnorm,
    args=c(0,1),
    xlim=c(-1.5, 1.2),
    geom="area",
    fill="#84CA72",
    alpha=0.2
  ) +
  ggtitle("A Normal Distribution") +
  geom_hline(yintercept=0)+ 
  geom_segment(
    aes(
      x = 0,
      y = 0,
      xend = 0,
      yend = dnorm(0, 0, 1),
    ),
    color="#389912",
    linetype="dashed"
  ) +
  theme(
    panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
    plot.title = element_text(hjust = 0.5, size=24), panel.border = element_blank(),
    axis.title.x = element_text(
      margin = margin(t = -20),
      hjust=0.95,
      size=24
    ),
    axis.text.x = element_text(size=25)
  ) +
  annotate(
    geom='text',
    x=0,
    y=0.18,
    label=TeX("$P(a\\leq X\\leq b)$",
              output='character'),
    parse=TRUE,
    size = 8
  )

pnormal
```

]

- Recall that the probability $P(X\le b)$ is a cumulative probability.

- By the addition rule of probability, we have $P(a\le X\le b)=P(X\le b)-P(X\le a)$ and $P(X\ge b)=1-P(X\le b)$

- As a line segment has no area, we have $P(X\le a)=P(X< a)$ as well as $P(X\ge b)=P(X>b)$

---

## Normal Distribution (1/5)

- A **normal distribution** has a **density function** $f(x)=\frac{1}{\sqrt{2\pi \sigma^2}}e^{-\frac{(x-\mu)^2}{2\sigma^2}},$
where $\mu$ is the mean, $\sigma$ is the standard deviation, $\pi\approx `r round(pi, 5)`$ and $e\approx `r round(exp(1), 5)`$. The graph of $f$ is called a **normal curve**.

- We write $X\sim \mathcal{N}(\mu, \sigma^2)$ for a normal random variable $X$ with the mean $\mu$ and the standard deviation $\sigma$.

- A normal distribution has the following properties:
  - *The mean, median, and mode are equal*.
  - The normal curve is *bell shaped and symmetric* with respect to the mean.
  - The *total area* under the curve and above the $x$-axis is $1$.
  - The normal curve *approaches, but never touches, the $x$-axis* as $x$ goes to $\pm\infty$.
  - Between $\mu-\sigma$ and $\mu+\sigma$, the graph *curves downward*. On the left side of $\mu-\sigma$ or the right side of $\mu+\sigma$, the graph *curves upward*.  A point at which the curve changes the direction of curving is called an **inflection point**.

---

## Normal Distribution (2/5)

.center[

```{r echo=FALSE, fig.width=7, fig.asp=0.5}
m1 <-0
m2 <- 1
m3 <- -1.5
mn <- c(m1, m2, m3)
pmean <- ggplot(
    data = data.frame(x = c(-4, 4)), aes(x)
  ) +
  stat_function(
    fun = dnorm,
    args = list(mean = m1, sd = 1),
    aes(colour = "B"),
    size = 1.25
  ) +
  stat_function(
    fun = dnorm,
    args = list(mean = m2, sd = 1),
    aes(colour = "G"),
    size = 1.25
  ) +
  stat_function(
    fun = dnorm,
    args = list(mean = m3, sd = 1),
    aes(colour = "R"),
    size = 1.25
  ) +
  xlab("") +
  ylab("") +
  scale_y_continuous(breaks=NULL) +
  theme_bw() +
  ggtitle(TeX('Normal Curves with $\\sigma = 1$')) +
  geom_hline(yintercept=0)+
  geom_segment(
    aes(
      x = m1,
      y = 0,
      xend = m1,
      yend = dnorm(m1, m1, 1),
    ),
    color="blue",
    linetype="dashed"
  ) +
  geom_segment(
    aes(
      x = m2,
      y = 0,
      xend = m2,
      yend = dnorm(m2, m2, 1),
    ),
    color="red",
    linetype="dashed"
  ) +
  geom_segment(
    aes(
      x = m3,
      y = 0,
      xend = m3,
      yend = dnorm(m3, m3, 1),
    ),
    color="red",
    linetype="dashed"
  ) +
  theme(
    panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
    plot.title = element_text(hjust = 0.5, size=24), panel.border = element_blank(),
    axis.title.x = element_text(
      margin = margin(t = -20),
      hjust=0.95,
      size=24
    ),
    axis.text.x = element_text(size=25)
  ) +
    scale_colour_manual(
    "",
    values= c("B"="blue", "R"="red", "G"="green"),
    labels = lapply(sprintf('$\\mu = %.1f$', mn), TeX)
  )

pmean
```
```{r echo=FALSE, fig.width=7, fig.asp=0.5}
sd1 <- 0.6
sd2 <- 1.5
sd3 <- 2
sd <- c(sd1, sd2, sd3)
psd <- ggplot(
    data = data.frame(x = c(-4, 4), y=c(0, 1)), aes(x, y)
  ) +
  stat_function(
    fun = dnorm,
    args = list(mean = 0, sd = sd1),
    aes(color="B"),
    size = 1.25
  ) +
  stat_function(
    fun = dnorm,
    args = list(mean = 0, sd = sd2),
    aes(color="G"),
    size = 1.25
  ) +
  stat_function(
    fun = dnorm,
    args = list(mean = 0, sd = sd3),
    aes(color="R"),
    size = 1.25
  ) +
  xlab("") +
  ylab("") +
  scale_y_continuous(breaks=NULL) +
  theme_bw() +
  ggtitle(TeX('Normal Curves with $\\mu = 0$')) +
  geom_hline(yintercept=0)+
  geom_segment(
    aes(
      x = 0,
      y = 0,
      xend = 0,
      yend = dnorm(0, 0, sd1),
    ),
    color="blue",
    linetype="dashed"
  ) +
  theme(
    panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
    plot.title = element_text(hjust = 0.5, size=24), panel.border = element_blank(),
    axis.title.x = element_text(
      margin = margin(t = -20),
      hjust=0.95,
      size=24
    ),
    axis.text.x = element_text(size=25)
  ) +
  scale_colour_manual(
    "",
    values= c("B"="blue", "R"="red", "G"="green"),
    labels = lapply(
          sprintf('$\\sigma = %.1f$', sd),
          TeX
      )
  )

psd
```

]

.center[

```{r echo=FALSE, fig.width=10, fig.asp=0.35}
sigma <- 1.5
br <- c(-3,-2,-1,0,1,2,3)*sigma
lb <- c(0.683, 0.954, 0.997)

ms <- ggplot(
  data = data.frame(x = c(-6, 6)), aes(x)
) +
  stat_function(
    fun = dnorm,
    args = list(mean = 0, sd = 1.5),
    color="blue",
    size = 1.25
  ) +
  xlab("") +
  ylab("") +
  scale_y_continuous(breaks=NULL) +
  scale_x_continuous(
    breaks=br,
    labels=c(
      quote(mu-3*sigma),
      quote(mu-2*sigma),
      quote(mu-sigma),
      quote(mu),
      quote(mu+sigma),
      quote(mu+2*sigma),
      quote(mu+3*sigma)
    )
  ) +
  theme_bw() +
  ggtitle(
    TeX('Area Under a Normal Curve but over $\\[ \\mu - n\\sigma, \\mu + n\\sigma \\]$')
  ) +
  geom_hline(yintercept=0) +
  theme(
    panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
    plot.title = element_text(hjust = 0.5, size=24), panel.border = element_blank(),
    axis.title.x = element_text(
      margin = margin(t = -20),
      hjust=0.95,
      size=24
    ),
    axis.text.x = element_text(size=20)
  )

for (i in br) {
  if( i != 0){
    ms <- ms + geom_segment(
      x = i,
      y = 0,
      xend = i,
      yend = dnorm(sigma, 0, sigma)+0.1+0.06*abs(i)/sigma,
      color="blue"
    )
  }
}

ms <- ms + geom_segment(
    aes(
      x = 0,
    y = 0,
    xend = 0,
    yend = dnorm(0, 0, sigma)
    ),
    color="blue"
)

ms <- ms +
  geom_point(
    aes(x=sigma, y=dnorm(sigma, 0, sigma)),
    size=3
  ) +
  geom_point(
    aes(x=-sigma, y=dnorm(-sigma, 0, sigma)),
    size=3
  )

ms <- ms + annotate(
  "segment",
  x = 2*sigma,
  xend = sigma,
  y = 0.23,
  yend = dnorm(sigma, 0, sigma),
  colour = "purple",
  alpha=0.6,
  arrow=arrow()
)+
  annotate(
    "segment",
    x = 2*sigma,
    xend = -sigma,
    y = 0.23,
    yend = dnorm(-sigma, 0, sigma),
    colour = "purple",
    alpha=0.6,
    arrow=arrow()
  ) +
  annotate(
    geom='text',
    x=3.1,
    y=0.24,
    label="Inflection Points",
    color = "purple",
    size=6
  )


ms <- ms +
  geom_segment(
    aes(
      x=-sigma,
      y=dnorm(sigma, 0, sigma)+0.1+0.06,
      xend=sigma,
      yend=dnorm(sigma, 0, sigma)+0.1+0.06
    ),
    color="blue",
  ) +
  geom_label(
      x=0,
      y=dnorm(sigma, 0, sigma)+0.1+0.06,
    label=lb[1]
  )

ms <- ms +
  geom_segment(
    aes(
      x=-2*sigma,
      y=dnorm(sigma, 0, sigma)+0.1+0.06*2,
      xend=2*sigma,
      yend=dnorm(sigma, 0, sigma)+0.1+0.06*2
    ),
    color="blue"
  ) +
  geom_label(
    x=0,
    y=dnorm(sigma, 0, sigma)+0.1+0.06*2,
    label=lb[2]
  )

ms <- ms +
  geom_segment(
    aes(
      x=-3*sigma,
      y=dnorm(sigma, 0, sigma)+0.1+0.06*3,
      xend=3*sigma,
      yend=dnorm(sigma, 0, sigma)+0.1+0.06*3
    ),
    color="blue",
  ) +
  geom_label(
    x=0,
    y=dnorm(sigma, 0, sigma)+0.1+0.06*3,
    label=lb[3]
  )

ms

```

]

---

## Normal Distribution (3/5)

- **Example:** Suppose that foot length of a randomly chosen adult male is a normal random variable with mean $\mu=11$ and standard deviation $\sigma=1.5$.
  - How likely is a male's foot length to be smaller than 9.5 inches
  - How likely is a male's foot length to be bigger than 8 inches

  **Solution:** Let's first sketch the normal curve.
.center[

```{r echo=FALSE, fig.width=6, fig.asp=0.25}
footsigma <- 1.5
footmu <- 11
footbr <- c(-2, -1, 0, 1, 2)*footsigma+footmu

foot <- ggplot(
  data = data.frame(x = c(7, 16)), aes(x)
) +
  stat_function(
    fun = dnorm,
    args = list(mean = footmu, sd = footsigma),
    color="blue",
    size = 1.25
  ) +
  stat_function(
    fun=dnorm,
    args=c(footmu,footsigma),
    xlim=c(6, 9.5),
    geom="area",
    fill="#5185ed",
    alpha=0.2
  ) +
  xlab("") +
  ylab("") +
  scale_y_continuous(breaks=NULL) +
  scale_x_continuous(
    breaks=footbr,
    labels=footbr
  ) +
  theme_bw() +
  geom_hline(yintercept=-0.01) +
  theme(
    panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
    plot.title = element_text(hjust = 0.5, size=24), panel.border = element_blank(),
    axis.title.x = element_text(
    margin = margin(t = -20),
    hjust=0.95,
    size=24
    ),
    axis.text.x = element_text(size=20)
  )

for (i in -1:1) {
    foot <- foot + geom_segment(
      x = i*footsigma+footmu,
      y = 0,
      xend = i*footsigma+footmu,
      yend = dnorm(i*footsigma+footmu, footmu, footsigma),
      color="blue"
    )
}

foot
 ```
```{r echo=FALSE, fig.width=6, fig.asp=0.25}
footsigma <- 1.5
footmu <- 11
footbr <- c(-2, -1, 0, 1, 2)*footsigma+footmu

foot2 <- ggplot(
  data = data.frame(x = c(7, 16)), aes(x)
) +
  stat_function(
    fun = dnorm,
    args = list(mean = footmu, sd = footsigma),
    color="blue",
    size = 1.25
  ) +
  stat_function(
    fun=dnorm,
    args=c(footmu,footsigma),
    xlim=c(8, 16),
    geom="area",
    fill="#5185ed",
    alpha=0.2
  ) +
  xlab("") +
  ylab("") +
  scale_y_continuous(breaks=NULL) +
  scale_x_continuous(
    breaks=footbr,
    labels=footbr
  ) +
  theme_bw() +
  geom_hline(yintercept=-0.01) +
  theme(
    panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
    plot.title = element_text(hjust = 0.5, size=24), panel.border = element_blank(),
    axis.title.x = element_text(
      margin = margin(t = -20),
      hjust=0.95,
      size=24
    ),
    axis.text.x = element_text(size=20)
  )

for (i in c(-2, 0, 2) ) {
    foot2 <- foot2 + geom_segment(
      x = i*footsigma+footmu,
      y = 0,
      xend = i*footsigma+footmu,
      yend = dnorm(i*footsigma+footmu, footmu, footsigma),
      color="blue"
    )
}

foot2
 ``` 
] 
  Note that $9.5=11-1.5=\mu-\sigma$. The probability $$\scriptstyle P(X<9.5)=\frac12(1-P(9.5<X<12.5))\approx\frac12(1-0.683)=0.1585.$$
  Note that $8=11-2\cdot 1.5=\mu-2\sigma$. The probability $$\scriptstyle P(X>8)=(1-P(X<8))=1-(0.5-0.5\cdot P(8<X<14))=0.5+0.5\cdot 0.954=0.977.$$

---

## Normal Distribution (4/5)

- A normal distribution is called a **standard normal distribution** if the mean is $\mu=0$ and the standard deviation is $\sigma=1$.

- A random normal variable can be **standardized** by the following formula
  $$z=\frac{x-\mu}{\sigma}.$$
  We call the value $z$ the ** $Z$-score** of $x$.

- In Excel, the $Z$-score of $x$ can be calculated by the function `STANDARDIZE()`.
  
- Standardization does not change probability:
  $$P(a<X<b)=P\left(\frac{a-\mu}{\sigma}< Z < \frac{b-\mu}{\sigma}\right).$$

- The probability $P(X< x)$ can be calculated by the function `NORM.DIST()` or the standard normal distribution table.

---

## Normal Distribution (5/5)

- **Example:** The heights of 25-year-old women in a certain region are approximately normally distributed with mean 62 inches and standard deviation 4 inches. Find the probability that a randomly selected 25-year-old woman is nore than 67 inches tall.

 **Solution:** Let's first sketch the normal curve.
.pull-left[

```{r echo=FALSE, fig.width=6, fig.asp=0.25}

heightsigma <- 5
heightmu <- 62
heightbr=c(62, 67)

height <- ggplot(
  data = data.frame(x=c(52, 72) ), aes(x)
) +
  stat_function(
    fun = dnorm,
    args = list(mean = heightmu, sd = heightsigma),
    color="blue",
    size = 1.25
  ) +
  stat_function(
    fun=dnorm,
    args=c(heightmu,heightsigma),
    xlim=c(67, 72),
    geom="area",
    fill="#5185ed",
    alpha=0.2
  ) +
  xlab("") +
  ylab("") +
  scale_y_continuous(breaks=NULL) +
  scale_x_continuous(
    breaks=heightbr,
    labels=heightbr
  ) +
  theme_bw() +
  geom_hline(yintercept=0) +
  theme(
    panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
    plot.title = element_text(hjust = 0.5, size=24), panel.border = element_blank(),
    axis.title.x = element_text(
    margin = margin(t = -20),
    hjust=0.95,
    size=24
    ),
    axis.text.x = element_text(size=20)
  ) +
  geom_segment(
      x = heightmu,
      y = 0,
      xend = heightmu,
      yend = dnorm(heightmu, heightmu, heightsigma),
      color="blue"
  ) +
  geom_segment(
      x = 67,
      y = 0,
      xend = 67,
      yend = dnorm(67, heightmu, heightsigma),
      color="blue"
  )

height
 ```
]
.pull-right[.center[Area to the left of the Z-score]

| Z    | 0.04   | 0.05   | 0.06   |
| ---- | ------ | ------ | ------ |
| 1.2  | 0.8925 | .red[0.8944] | 0.8962 |
]
<br>
The probaiblity is $P(X>67)=1-P(X<67)$. To calculate $P(X<67)$:
<br>
  **Method 1:** Use the standard normal distribution table: First find the $Z$-score: $z=\frac{67-62}{4}=1.25$; Then $P(Z<1.25)\approx 0.8944$ is the value located in the row of 1.2 (first two digits of the $z$-score) and the column 0.05 (the last digit of the $z$-score).
<br>
  **Method 2:** Use Excel:  function `NORM.DIST()`: $P(X<67)=$ `NORM.DIST(67, 62, 4, TRUE)` $\approx 0.8944$.
  <br>
  Then $P(X>67)\approx 1-0.8944=0.1056$.

---
class: center middle

# Exercises

---

## Exercises on Expected Value and Standard Deviation

- **Exercise:** Five thousand lottery tickets are sold for $1 each. One ticket will win $1,000, two tickets will win $500 each, and five tickets will win $200 each. Find the expect gain if one ticket was bought.

.pull-left[

- **Exercise:** A company tracks the number of sales new employees make each day during a 100-day probationary period. The results for one new employee are shown at the right.

  1. Find the probability of each outcome.
  2. Construct a probability distribution table.
  3. Find the mean of the probability distribution.
  4. Find the variance and standard deviation.
]<br>
.pull-right[

| Sales per day $x$ | Number of days $f$ |
|-------------------|--------------------|
| 0                 | 16                 |
| 1                 | 19                 |
| 2                 | 15                 |
| 3                 | 21                 |
| 4                 | 9                  |
| 5                 | 10                 |
| 6                 | 8                  |
| 7                 | 2                  |
]

---

## Exercises on Binomial and Normal Distribution

- **Exercise:** Let $X$ be a binomial random variable with parameters $n = 5$, $p=0.2$. Find the probabilities
  $\text{1.}\,\, P(X=3) \quad \text{2.}\,\, P(X<3)\quad \text{3.}\,\, P(X>3).$

- **Exercise:** Let $Z$ be a standard normal random variable. Find the probabilities:
  $\text{1.}\,\, P(Z<1.58)\quad \text{2.}\,\,  P(-0.6<Z<1.67)\quad \text{3.}\,\, P(Z>0.19).$

- **Exercise:** Let $X$ be a normal random variable with $\mu=5$ and $\sigma=2$. Find the probabilities:
  $\text{1.}\,\,  P(-2<X<8)\quad \text{2.}\,\, P(X>-1) \quad \text{3.}\,\, P(X<4).$

- **Exercise:** A type of surgery has a 90% chance of success. The surgery is performed on three patients. Find the probability of the surgery being successful on exactly two patients.

- **Exercise:** Exercise 17, 19 in [The Binomial Distribution](https://saylordotorg.github.io/text_introductory-statistics/s08-03-the-binomial-distribution.html)  in Introductory Statistics

- **Exercise:** Exercises in [Normal Random Variables (5 of 6)](https://courses.lumenlearning.com/wmopen-concepts-statistics/chapter/introduction-to-normal-random-variables-5-of-6/) and [Normal Random Variables (6 of 6)](https://courses.lumenlearning.com/wmopen-concepts-statistics/chapter/introduction-to-normal-random-variables-6-of-6/) in Concepts in Statistics.

---
class: center middle

# Quiz

---

## Quiz 6

Let $Z$ be a normal random variable with $\mu=0$ and $\sigma=1$. Let $X$ be a normal random varialbe with $\mu=4.3$ and $\sigma=1.7$.

- Sketch the normal curves for the two random variable.

- Determine the values $P(Z>1) + P(X<6)$ and explain how do you find the value.

---
class: center middle

# Lab Instructions in Excel

---

## Excel Functions for Binomial and Normal Distributions

- Let $X$ be a binomial random variable with parameters $n$ and $p$, that is $X\sim B(n, p)$. In Excel, $P(X=x)$ is given by `BINORM.DIST(x, n, p, FALSE)` and $P(X\le x)$ is given by `BINORM.DIST(x, n, p, true)`. You may click input function $f_x$ and then search `binorm` to find the function.

- Let $Z$ be a standard normal random varaible. In Excel, $P(Z<z)$ is given by `NORM.S.DIST(z, TRUE)`.

- Let $X$ be a normal random variable with mean $\mu$ and standard deviation $\sigma$, that is $X\sim \mathcal{N}(\mu, \sigma^2)$. In Excel, $P(X<x)$ is given by `NORM.DIST(x, mu, sigma, TRUE)`.

- When a cumulative probability $p=P(X<x)$ of a normal random variable $X$ is given, we can find $x$ using `NORM.INV(p, mu, sigma)`.

- When a cumulative probability $p=P(Z<z)$ of a standard normal random variable $Z$ is given, we can find $z$ using `NORM.S.INV(p)`.