---
title: "Sampling Distributions"
subtitle: "MA336 Statistics<br/><br/>"
author: "Fei Ye <br/><br/> Department of Mathematics and Computer Science<br/><br/>"
date: '`r format(Sys.time(), "%B, %Y")`'
output:
  xaringan::moon_reader:
    css: [default, default-fonts, "Assets/css/custom.css"]
    mathjax: "https://cdnjs.cloudflare.com/ajax/libs/mathjax/2.7.4/MathJax.js?config=TeX-MML-AM_HTMLorMML-full"
    nature:
      ratio: '16:9' # alternatives '16:9' or '4:3' or others e.g. 13:9
      highlightLines: true
      highlightStyle: github
      countIncrementalSlides: false  
      beforeInit: ["Assets/js/resize.js"]
---

```{r, echo=FALSE}
knitr::opts_chunk$set(comment="#", fig.retina=2, crayon.enabled = TRUE, cache=TRUE, dpi=300)
set.seed(15)
library(stats)
library(kableExtra)
library(knitr)
library(formattable)
library(ggplot2)
library(ggthemes)
library(ggExtra)
library(dplyr,warn.conflicts = FALSE)
library(animation)
library(gifski)
library(latex2exp)
```

# Textbooks

- [Concepts in Statistics](https://courses.lumenlearning.com/wmopen-concepts-statistics/)

- [Introductory Statistics](https://open.umn.edu/opentextbooks/textbooks/introductory-statistics)


---
class: center middle

# Learning Goals

---

## Learning Goals for Sampling Distribution

- Demonstrate understanding of the sampling distribution of a statistic.

- Explain how the central limit theorem applies in inference.

- Determine whether a sampling distribution is approximately a normal distribution.

- Calculate key characteristics (mean, standard error) of the sampling distribution of a statistic.

- Estimate the probability of an event using the sampling distribution.

---
class: center middle

# Central Limit Theorem

---

## Sampling Distribution

- When using sample statistics to estimate population parameter, there will be a chance error
  $$\text{Statistic}=\text{Parameter}+\text{Chance Error}.$$

- To understand the chance error, we need to know how sample statistics distribute. Consider samples of the same size $n$ randomly chosen from the population with replacement.  The distribution of sample statistics is called a **sampling distribution**.

  **Example:** Sampling distributions of sample means.

  - [Sampling Distribution of the Sample Mean (discrete population)](https://istats.shinyapps.io/SampDist_discrete/)
  
  - [Sampling Distribution of the Sample Mean (continuous population)](https://istats.shinyapps.io/sampdist_cont/)

- The sampling distribution varies as the sample size changes. In the above examples, the standard deviation of the sampling distribution changes when $n$ changes.

- The standard deviation of a sampling distribution is also called the **standard error**.

---

## Central Limit Theorem (1/3)

- **The Central Limit Theorem:**

  As the sample size $n$ increases, the sampling distribution of the sample mean, from a population with the mean $\mu$ and the standard deviation $\sigma$, will approach to a normal distribution with the mean $\mu_{\bar{X}}=\mu$ and the standard deviation $\sigma_{\bar{X}}=\dfrac{\sigma}{\sqrt{n}}$.

- **Two Practical Facts:**

  - For most distributions (not highly skewed), when sample size $n>30$, the sampling distribution of the sample mean $\bar{X}$ can be approximated reasonably well by a normal distribution. The larger the sample size, the better the approximation will be.

  - When the population is normally distributed, the sampling distribution of the sample means will be normally distributed for any sample size.

- **Remark:** In terms of standardization, the central limit theorem says that the random variable $\bar{Z}=\dfrac{\bar{x}-\mu}{\sigma/\sqrt{n}}$ has an approximately standard normal distribution.

???
- [See the discussion on intuitive explanation.](https://stats.stackexchange.com/questions/3734/what-intuitive-explanation-is-there-for-the-central-limit-theorem/3904#3904)

- If the population distribution is highly skewed, relying on CLT can be risky.

---

## Central Limit Theorem (2/3)

**Example:** Randomly draw samples of size 2 with replacement from the numbers 1, 3, 4.

- List all possible samples and calculate the mean of each sample.
- Find the mean, and standard deviation of the sample means.
- Find the mean, and standard deviation of the population.

**Solution:** Using the Excel function `AVERAGE()`, we may find means of samples and means of sample means.

Using the Excel function `STDEV.P()`, we may find the standard deviation of the population and the standard deviation of sample means.

.center[

.left-column[
  | $\color{red}{\mu}$ | $\color{red}{\sigma}$ | $\color{blue}{\mu_{\bar{X}}}$ | $\color{blue}{\sigma_{\bar{X}}}$ |
  | --------- | ----- | ----- | ----- |
  | .red[2.7]  | .red[1.25] |.blue[2.7]|.blue[0.88]|
]

.right-col[
  | sample   | (1,1) | (1,3) | (1,4) | (3,1) | (3,3) | (3,4) | (4,1) | (4,3) | (4,4) |
  | --------- | ----- | ----- | ----- | ----- | ----- | ----- | ----- | ----- | ----- |
  | $\bar{X}$ | 1     | 2     | 2.5   | 2     | 3     | 3.5   | 2.5   | 3.5   | 4     |
]

]

It can be verified that $\mu_{\bar{X}}=\mu$ and $\sigma/\sqrt{n}=1.25/\sqrt{2}\approx 0.88=\sigma_{\bar{X}}$.

---

## Central Limit Theorem (3/3)

**Example:(continued)** The following are the distribution of the population and the distribution of sample means.

.pull-left[
.center[

```{r echo=FALSE, results = 'asis', fig.width=5, fig.asp=0.8}

pop <- c(1,3,4)

popdf <- data.frame(pop)

colnames(popdf) <- c("x")

popplot <- ggplot(popdf, aes(x=x), dpi=600) +
  geom_histogram(aes(y=..density..),bins=3,fill="lightblue", col="white") +
  theme_bw(base_size = 12) +
geom_vline(aes(xintercept=mean(x)),       color="blue", linetype="dashed", size=1)+
labs(x = TeX('$x$'), y = "rel. freq.")

popplot
```

]
]

.pull-right[
.center[

```{r echo=FALSE, results = 'asis', fig.width=5, fig.asp=0.8}

samplemean <- c(1,2,2.5,2,3,3.5,2.5,3.5,4)

smdf <- data.frame(samplemean)

colnames(smdf) <- c("xbar")

clt <- ggplot(smdf, aes(x=xbar), dpi=600) +
  geom_histogram(aes(y=..density..), bins=4,fill="lightblue", col="white") +
  theme_bw(base_size = 12) +
geom_vline(aes(xintercept=mean(xbar)),       color="blue", linetype="dashed", size=1)+
labs(x = TeX('$\\bar{X}$'), y = "rel. freq.")

clt +  stat_function( fun = dnorm, args = list(mean=mean(samplemean), sd=sd(samplemean)),color="gold", xlim = c(0.5, 4.5) )
```

]
]

---
class: center middle

# Sampling Distributions for Mean and Proportion

---

## Sampling Distribution of the Sample Mean (1/2)

**Example:** Suppose the mean length of time that a caller is placed on hold when telephoning a customer service center is 23.8 seconds, with standard deviation 4.6 seconds. Find the probability that the mean length of time on hold in a sample of 1,000 calls will be within 0.5 second of the population mean.

**Solution:**

  Since the sample size $n=1000>30$ is large enough, by the Central Limit Theorem, we know the mean length of time is approximately normally distributed.

  The mean of the sampling distribution is $\mu_{\bar{X}}=\mu=23.8$.

  The standard deviation of the sampling distribution is $\mu_{\bar{X}}=\dfrac{\sigma}{\sqrt{n}}=\dfrac{4.6}{\sqrt{1000}}\approx 0.15.$

  By the Excel function `norm.dist(xbar, mean, sd, true)`, the probability is calculated as
  $$\begin{align}P(23.8-0.5<\bar{X}<23.8+0.5)= & P(\bar{X}<24.3)-P(\bar{X}<23.3)\\ \approx & 0.9997-0.0003=0.9994.\end{align}$$

---

## Sampling Distribution of the Sample Mean (2/2)

**Example:** Suppose speeds of vehicles on a particular stretch of roadway are normally distributed with mean 36.6 mph and standard deviation 1.7 mph.

- Find the probability that the speed $X$ of a randomly selected vehicle is between 35 and 40 mph.
- Find the probability that the mean speed $\bar{X}$ of 10 randomly selected vehicles is between 35 and 40 mph.

**Solution:** In this example, the population is normally distributed. So the sampling distribution of the sample mean is always normally distributed. For calculation, the Excel function `norm.dist(X, mean, sd, true)` will be used.

As $\mu=36.6$ and $\sigma=1.7$, the probability that the speed of a vehicle is between 35 and 40 is $P(35< X< 40)=P(X< 40)-P(X<35)\approx 0.9772-0.1733=0.8039$

The mean of the sampling distribution is $\mu_{\bar{x}}=\mu=36.6$. The standard deviation of the sampling distribution is $\sigma_{\bar{X}}=\sigma/\sqrt{n}=1.7/\sqrt{10}\approx 0.54.$ Then the probability is $P(35<\bar{X}< 40)=P(\bar{X}< 40)-P(\bar{X}<35)\approx 1-0.0015=0.9985.$

---

## Sampling Distribution of the Sample Proportion (1/3)

- When working with categorical variables, we often study the proportion of a data set.

- The proportion of a specific characteristic in a data set can be viewed as the mean of the data set by identifying the specific characteristic with 1 and others with $0$.

  **Example:** Consider the following data set
  .center[.red[1], 0, .red[1], .red[1], 0, 0, .red[1], 0, .red[1], .red[1]]
  Find proportion of .red[red numbers] and the mean of the data set.

  **Solution:** The proportion of red numbers is $\frac{6}{10}=0.6$. So is the mean: $\frac{6\cdot 1 + 4\cdot 0}{10}=0.6$.

- Consider a population consisting of 1s and 0s. Let $p$ be the proportion of 1s. Then standard deviation $$\sigma=\sqrt{(1-p)^2p+(0-p)^2(1-p)}=\sqrt{p(1-p)}.$$

---

## Sampling Distribution of the Sample Proportion (2/3)

- **Central Limit Theorem for Sample Proportion:** For large samples, the distribution of sample proportions $\hat{P}$ is approximately normal, with the mean $\mu_{\hat{P}}=p$ and standard deviation $\sigma_{\hat{P}}=\sqrt{\frac{p(1-p)}{n}}$, where $p$ is the population proportion.

- As a sample proportion is always between 0 and 1, and 99.7% of sample proportions lie within 3 standard deviation away, we require the sample size $n$ satisfying the following condition: the interval $\left[p-3\sqrt{\frac{p(1-p)}{n}}, p+3\sqrt{\frac{p(1-p)}{n}}\right]$ lies wholly in the interval $[0, 1]$.

- In practice, if $n$ satisfies the following two inequalities: $np\ge 10$ and $n(1-p)\ge 10$, then we consider $n$ is large enough for assuming that the sampling distribution of the sample proportion is approximately normal.

---

## Sampling Distribution of the Sample Proportion (3/3)

**Example:** Suppose that in a population of voters in a certain region 53% are in favor of a particular law. Nine hundred randomly selected voters are asked if they favor the law.

Find the probability that the sample proportion computed from a random sample of size 900 will be at least 2% above true population proportion.

**Solution:** We first verify that the sampling distribution is approximately normal.

Since $p=0.53$ and $n=900$, $np=900\cdot 0.53>10$ and $n(1-p)=900(1-0.53)>10$. By the central limit theorem, the sampling distribution is approximately normal.

The standard deviation of the sampling distribution is $\sigma_{\hat{P}}=\sqrt{\frac{0.53(1-0.53)}{900}}\approx 0.017$.

Then the probability that the random sample has a proportion at least 2% above 53% is
  $$P(\hat{P}>0.55)=1-P(\hat{P}\le 0.55)\approx 1-0.8803=0.1197.$$

---

## More Examples on Sampling Distributions

- Example 4 in [Section 6.2 The Sampling Distribution of the Sample Mean](https://saylordotorg.github.io/text_introductory-statistics/s10-02-the-sampling-distribution-of-t.html) in Introductory Statistics.

- Example 5 in [Section 6.2 The Sampling Distribution of the Sample Mean](https://saylordotorg.github.io/text_introductory-statistics/s10-02-the-sampling-distribution-of-t.html) in Introductory Statistics.

- Example 8 in [Section 6.3 The Sample Proportion](https://saylordotorg.github.io/text_introductory-statistics/s10-03-the-sample-proportion.html) in Introductory Statistics.

---
class: center middle

# Exercise

---

## Exercises on Sampling Distributions

- Exercise 3 in [Section 6.2 The Sampling Distribution of the Sample Mean](https://saylordotorg.github.io/text_introductory-statistics/s10-02-the-sampling-distribution-of-t.html) in Introductory Statistics.

- Exercise 6 in [Section 6.2 The Sampling Distribution of the Sample Mean](https://saylordotorg.github.io/text_introductory-statistics/s10-02-the-sampling-distribution-of-t.html) in Introductory Statistics.

- Exercise 15 in [Section 6.2 The Sampling Distribution of the Sample Mean](https://saylordotorg.github.io/text_introductory-statistics/s10-02-the-sampling-distribution-of-t.html) in Introductory Statistics.

- Exercise 9 in [Section 6.3 The Sample Proportion](https://saylordotorg.github.io/text_introductory-statistics/s10-03-the-sample-proportion.html) in Introductory Statistics.

- Exercise 13 in [Section 6.3 The Sample Proportion](https://saylordotorg.github.io/text_introductory-statistics/s10-03-the-sample-proportion.html) in Introductory Statistics.

- Exercise 17 in [Section 6.3 The Sample Proportion](https://saylordotorg.github.io/text_introductory-statistics/s10-03-the-sample-proportion.html) in Introductory Statistics.

---
class: center middle

# Quiz

---

## Quiz 7

1. State the Central Limit Theorem for Proportion.

2. Determine whether the given sampling distribution is approximately normal. **Please explain!**

   1. The sampling distribution of means of samples of size 50.

   2. The sampling distribution of proportions of samples of size 50.

---
class: center middle

# Lab Instructions in Excel

---

## The `NORM.DIST()` Function

- Let $X$ be a normal random variable with mean $\mu$ and standard deviation $\sigma$, that is $X\sim \mathcal{N}(\mu, \sigma^2)$. In Excel, $P(X<x)$ is given by `norm.dist(x, mean, sd, true)`.

- The $Z$-score of a value $x$ of a Random variable can be calculated using the Excel function `standardize(x, mean, sd)`.

- Let $Z$ be a standard normal random variable. The probability $P(Z<z)$ can be calculated using the Excel function `norm.s.dist(z)`