---
title: "Confidence Intervals for Proportion"
subtitle: "MA336 Statistics<br/><br/>"
author: "Fei Ye <br/><br/> Department of Mathematics and Computer Science<br/><br/>"
date: '`r format(Sys.time(), "%B %Y")`'
output:
  xaringan::moon_reader:
    self_contained: yes
    mode: selfcontained
    css: [default, default-fonts, "Assets/css/custom.css"]
    mathjax: "https://cdnjs.cloudflare.com/ajax/libs/mathjax/2.7.4/MathJax.js?config=TeX-MML-AM_HTMLorMML-full"
    nature:
      ratio: '16:9' # alternatives '16:9' or '4:3' or others e.g. 13:9
      highlightStyle: github
      countIncrementalSlides: false  
      beforeInit: ["Assets/js/resize.js"]
---

```{r, echo=FALSE, warning=FALSE}
knitr::opts_chunk$set(comment="#", fig.retina=2, crayon.enabled = TRUE, dpi=600)
set.seed(15)
library(stats)
library(kableExtra)
library(knitr)
library(formattable)
library(ggplot2)
library(ggthemes)
library(ggExtra)
library(dplyr,warn.conflicts = FALSE)
library(animation)
library(gifski)
library(latex2exp)
# library(learnr)
xaringanExtra::use_fit_screen()
```

```{css, echo=F}
.remark-slide table, .remark-slide table thead th {
    border-top: 0px;
    border-bottom: 0px;
}
.remark-slide thead, .remark-slide tr:nth-child(even){
  background-color: white;
}
table{
  border-collapse: collapse;
}
.remark-slide thead:empty {
  display: none;
}
```

## Learning Goals for Confidence Intervals

- Construct and interpret a confidence intervals for one population proportion.

- Describe how the following will affect the width of the confidence interval:

  - increasing the sample size;

  - increasing the confidence level.

---

## Confidence Interval for a Proportion (1/2)

- Recall that the standard error of sample proportions is $\sigma_{\hat{P}}=\sqrt{\frac{p(1-p)}{n}}$, where $n$ is the sample size and $p$ is the population proportion. As a consequence, when estimating the population proportion $p$, we only have a point estimate $\hat{p}$ (phat) to use. For the standard error, we use the estimation
  $$\sigma_{\hat{p}}\approx\hat{\sigma}_{\hat{p}}=\sqrt{\dfrac{\hat{p}(1-\hat{p})}{n}}.$$

- Based on the central limit theorem, when $n$ is large enough, at the $100(1-\alpha)\%$ level, the margin of error for $p$ is defined as
  $$E=z_{\alpha/2}\sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$$
  In Excel, $z_{\alpha/2}$=`NORM.S.INV((1 + confidence level)/2)`. The marginal error can also be obtained by
  `CONFIDENCE.NORM(1-confidence level, SQRT(phat*(1-phat)/n, n)`.

---

## Confidence Interval for a Proportion (2/2)

- The confidence interval for $p$ is defined by
  $$[\hat{p}-E,\hat{p}+E]=\left[\hat{p}-z_{\alpha/2}\sqrt{\frac{\hat{p}(1-\hat{p})}{n}}, \hat{p}+z_{\alpha/2}\sqrt{\frac{\hat{p}(1-\hat{p})}{n}}\right],$$
  where [the critical value $z_{\alpha/2}$ satisfies that $P(Z< z_{\alpha/2})=1-\alpha/2$](https://saylordotorg.github.io/text_introductory-statistics/s11-01-large-sample-estimation-of-a-p.html) for the standard normal variable $Z$.

- In practical, the sample size $n$ is considered large enough if $n\hat{p}\ge 10$ and $n(1-\hat{p})\ge 10$.

- The above defined confidence interval is known as the [normal approximation (or Wald's) confidence interval](https://en.wikipedia.org/wiki/Binomial_proportion_confidence_interval).
  It is popular in introductory statistics books. However, it is unreliable when the sample size is small or the sample proportion is close to 0 or 1.
  Indeed, if the sample proportion is 0 or 1, the confidence interval defined here will have zero length.

???
By the central limit theorem, the random variable $\hat{p}$ is normal distributed. The chance that $p\in \left[\hat{p}-z_{\alpha/2}\sqrt{\frac{\hat{p}(1-\hat{p})}{n}}, \hat{p}+z_{\alpha/2}\sqrt{\frac{\hat{p}(1-\hat{p})}{n}}\right]$ is the same as the chance that $\hat{p}\in \left[p-z_{\alpha/2}\sqrt{\frac{p(1-p)}{n}}, p+z_{\alpha/2}\sqrt{\frac{p(1-p)}{n}}\right]$. That shows $z_{\alpha/2}$ satisfying
$$P(-z_{\alpha/2}<\dfrac{\hat{p}-p}{\sqrt{\frac{p(1-p)}{n}}<z_{\alpha/2})=1-\alpha.$$

---

## Example: Estimating the Proportion of Students Taking Busses (1/2)

In a random sample of 100 students in college, 65 said that they come to college by bus.

1. Give a point estimate of the proportion of all students who come to college by bus.

2. Construct a 99% confidence interval for that proportion.

--

**Solution:** A good point estimate would be a sample proportion. Here the sample proportion is $\hat{p}=65/100=0.65$.

As $n\hat{p}=100\cdot 0.65=65>10$ and $n(1-\hat{p})=100\cdot 0.35=35>10$, which implies the sample is large enough, approximately the standard error is
$$\hat{\sigma}_{\hat{P}}=\sqrt{\frac{\hat{p}(1-\hat{p})}{n}}=\sqrt{\frac{0.65(1-0.65)}{100}}\approx`r round(sqrt(0.65*(1-0.65)/100), 3)`.$$

---

## Example: Estimating the Proportion of Students Taking Busses (2/2)

**Solution: (Continued)** At 99% level of confidence, the value $1-\alpha/2=(1+\text{confidence level})/2=(1+0.99)/2=0.995$. The critical value $z_{\alpha/2}$ is determined by the equation $P(Z<z_{\alpha/2})=0.9955$. Using the Excel function `NORM.S.INV(0.995)`, we find the critical value \\(z_{\alpha/2}\approx 2.576\\).

Thus the marginal error is
$$E=z_{\alpha/2}\cdot \hat{\sigma}_{\hat{P}}=`r round(qnorm(1-0.01/2,0,1),3)`\cdot `r round(sqrt(0.65*(1-0.65)/100), 3)`=`r busME=round(qnorm(1-0.01/2,0,1)*sqrt(0.65*(1-0.65)/100), 3); busME`,$$
and the confidence interval at 99% level is
$$[\hat{p}-E, \hat{p}+E]\approx [0.65-`r busME`, 0.65+`r busME`]=[`r 0.65- busME`, `r 0.65+busME`].$$

Conclusion: we are 99% confident that the proportion of all students at the college who take bus is in the interval $[`r 0.65- busME`, `r 0.65+busME`]$.

.footmark[
  **Note:** The marginal error can also be obtained by the Excel function
  `CONFIDENCE.NORM(1-0.99, SQRT(65/100*(1-65/100)/100), 100)`.
]

<!-- --- -->

<!-- ## Example: Estimating a Population Proportion (1/2) -->
<!-- 
Foothill College’s athletic department wants to calculate the proportion of students who have attended a women’s basketball game at the college. They use student email addresses, randomly choose 220 students, and email them. Of the 145 who responded, 22 had attended a women’s basketball game.

Calculate and interpret the approximate 90% confidence interval for the proportion of all Foothill College students who have attended a women’s basketball game. -->

<!-- -- -->
<!-- 
**Solution:** Although 220 students were surveyed, only 145 responded. So the sample consists of those 145 students. The sample proportion is $\hat{p}=\frac{22}{145}=`r phat=round(22/145,3);phat`$.

The estimated standard error is
$$\hat{\sigma}_{\hat{P}}=\sqrt{\frac{\hat{p}(1-\hat{p})}{n}}=\sqrt{\frac{`r phat`\cdot(1-`r phat`)}{145}}\approx `r SE=round(sqrt(phat*(1-phat)/145),3);SE`.$$ -->

<!-- --- -->

<!-- ## Example: Estimating a Population Proportion (2/2) -->

<!-- **Solution: (Continued)**
At the 90% confidence level, $\alpha=1-0.9=0.1$ and the critical value is $z_{\alpha/2}$=`NORM.S.INV(1-0.1/2)` $\approx `r zalpha=round(qnorm(1-0.1/2,0,1),3);zalpha`$.

Therefore, the marginal error is
$$E=z_{\alpha/2}\cdot\hat{\sigma}_{\hat{P}}\approx `r zalpha`\cdot`r SE`=`r ME=round(zalpha*SE,3); ME`,$$
and the confidence interval is
$$[\hat{p}-E, \hat{p}-E]\approx[`r phat`-`r ME`, `r phat`+`r ME`]=[`r phat - ME`, `r + ME`].$$

Conclusion: we are 90% confident that the proportion of all Foothill College students who have attended a women’s basketball game is between `r phat - ME` and `r phat + ME`. -->

<!-- .footmark[
  Source: [Estimating a Population Proportion in the book Concepts in Statistics](https://courses.lumenlearning.com/wmopen-concepts-statistics/chapter/estimating-a-population-proportion-1-of-3/)
]  -->

---

## Factors Affect the Width of Confidence Intervals

- The width of a confidence interval, equals twice the standard error, gives a measure of precision of the estimation.

- Recall, for population proportion and mean,
  $$\text{Marginal Error} = \text{Critical Value}\cdot \frac{\text{(estimated) Population SD}}{\sqrt{\text{Sample Size}}}$$

- The formula tells us the precision of a confidence interval is affected by the confidence level, the variability, and the sample size.
  
  - Larger the confidence levels give larger critical values and errors.
  
  - Populations (and samples) with more variability gives larger errors.

  - Larger sample sizes give smaller errors.

---

## Sample Size Determination

- In practice, we may desire a marginal error of $E$. With a fixed confidence level $100(1-\alpha)\%$, the larger the sample size the smaller the marginal error.

- When estimating population proportion, if we can produce a reasonable guess $\hat{p}$ for population proportion, then an appropriate minimum sample size for the study is determined by
  $$n=\left(\frac{z_{\alpha/2}}{{E}}\right)^2\cdot \hat{p}(1-\hat{p}).$$

- When estimating population mean, if we can produce a reasonable guess $\sigma$ for the population standard deviation, then an appropriate minimum sample size is given by
  $$n=\left(\dfrac{z_{\alpha/2}\cdot \sigma}{{E}}\right)^2.$$

---

## Example: Minimum Sample Size - Error in Proportion

Suppose you want to estimate the proportion of students at QCC who live in Queens. By surveying your classmates, you find around 70% live in Queens. Use this as a guess to determine how many students would need to be included in a random sample if you wanted the error of margin for a 95% confidence interval to be less than or equal to 2%.

**Solution:** We may use $\hat{p}=0.7$ as a reasonable guess for the population proportion.

At the 95% level, the critical value is $z_{\alpha/2}=$ `NORM.S.INV((1+0.95)/2)` $\approx 1.96$.

The marginal error is $E=0.02$.

Then the appropriate minimal sample size is determined by
$$n=\left(\frac{z_{\alpha/2}}{{E}}\right)^2\cdot \hat{p}(1-\hat{p})=(1.96/0.02)^2\cdot 0.7\cdot(1-0.7)=2016.84.$$

Since the sample size has to be an integer, to get a error no more than 2% at the level 95%,  the minimal sample size should be at least 2017.

---

## Example: Minimum Sample Size - Error in Mean

Find the minimum sample size necessary to construct a 99% confidence interval for the population mean with a margin of error $E =0.2$. Assume that the estimated population standard deviation is $\sigma=1.3$.

**Solution:** At the 99% level, the critical value $z_{\alpha/2}=$ `NORM.S.INV((1+0.99)/2)` $\approx 2.576$.

The desired marginal error is ${E}=0.2$.

The estimated population standard deviation is $\sigma=1.3$.

Then the minimal sample size is approximately
$$n=\left(\dfrac{z_{\alpha/2}\cdot \sigma}{{E}}\right)^2\approx (2.576\cdot 1.3/0.2)^2 \approx 280.4.$$

To get a error no more than 0.2 at the level 95%,  the minimal sample size should be at least 281.

---

## Practice: Confidence Interval of Proportion of Kids

`r knitr::include_url('https://www.myopenmath.com/embedq2.php?id=7347&seed=2020&showansafter', height='560px')`

---

## Practice: Confidence Intervals for Product Quality

To understand the reason for returned goods, the manager of a store examines the records on 40 products that were returned in the last year. Reasons were coded by 1 for “defective,” 2 for “unsatisfactory,” and 0 for all other reasons, with the results shown in the table.
```{r echo=FALSE}
returned <- c(0,0,0,0,2,0,0,0,0,0,2,0,0,0,0,0,0,0,0,0,0,0,0,1,0,0,0,0,2,0,2,0,0,0,0,0,0,2,0,0)
returned.df=matrix(returned,ncol=20,byrow=TRUE)
returned.df <- as.data.frame(returned.df)

returned.df[] <- mutate_all(returned.df, function(x) {
    cell_spec(x,
      color = spec_color(x, begin=0.4, end = 0.9,  direction = 1)
    )
  }
)

kable(returned.df, col.names = NULL, escape = F, align = "c", "html", booktabs = T) %>%
  kable_styling()
```

1. Give a point estimate of the proportion of all returns that are because of something wrong with the product, that is, either defective or performed unsatisfactorily.

2. Construct an 80% confidence interval for the proportion of all returns that are because of something wrong with the product.

---

## Practice: Sample Size for Mean with Given Error

`r knitr::include_url('https://www.myopenmath.com/embedq2.php?id=19799&seed=2020&showansafter', height='560px')`

---

## Practice: Sample Size for Proportion with Given Error

`r knitr::include_url('https://www.myopenmath.com/embedq2.php?id=7348&seed=2020&showansafter', height='560px')`

<!-- --- -->

<!-- ## Practice: Confidence Interval of Proportion Given Table -->

<!-- `r knitr::include_url('https://www.myopenmath.com/embedq2.php?id=213816&seed=2020&showansafter', height='560px')` -->

<!-- --- -->

<!-- ## Practice: Minimum Sample Size - Mean -->
<!-- 
A software engineer wishes to estimate, to within 5 seconds, the mean time that a new application takes to start up, with 95% confidence. Estimate the minimum size sample required if the standard deviation of start up times for similar software is 12 seconds.

.footmark[
 Source: [Exercise 7 in Section 7.4 in Introductory Statistics](https://saylordotorg.github.io/text_introductory-statistics/s11-04-sample-size-considerations.html)
] -->

<!-- --- -->

<!-- ## Practice: Minimum Sample Size - Proportion -->
<!-- 
The administration at a college wishes to estimate, to within two percentage points, the proportion of all its entering freshmen who graduate within four years, with 90% confidence. Estimate the minimum size sample required.

.footmark[
 Source: [Exercise 13 in Section 7.4 in Introductory Statistics](https://saylordotorg.github.io/text_introductory-statistics/s11-04-sample-size-considerations.html)
] -->

---
class: center middle

# Lab Instructions in Excel

---

## Excel Functions for Normal Distributions and Marginal Errors

- When a cumulative probability $p=P(Z<z)$ of a standard normal random variable $Z$ is given, we can find $z$ using `NORM.S.INV(p)`.

- If a sample of size $n$ has the proportion $\hat{p}$ and the sampling distribution is approximately normal, the marginal error for the proportion can be obtained by the Excel function
`CONFIDENCE.NORM(1-confidence level, SQRT(phat*(1-phat)), n)`
