---
title: "Discrete Random Variables"
subtitle: "MA336 Statistics<br/><br/>"
author: "Fei Ye <br/><br/> Department of Mathematics and Computer Science<br/><br/>"
date: '`r format(Sys.time(), "%B, %Y")`'
output:
  xaringan::moon_reader:
    self_contained: yes
    mode: selfcontained
    css: [default, default-fonts, "Assets/css/custom.css"]
    mathjax: "https://cdnjs.cloudflare.com/ajax/libs/mathjax/2.7.4/MathJax.js?config=TeX-MML-AM_HTMLorMML-full"
    nature:
      ratio: '16:9' # alternatives '16:9' or '4:3' or others e.g. 13:9
      highlightStyle: github
      countIncrementalSlides: false  
      beforeInit: ["Assets/js/resize.js"]
---

```{r, echo=FALSE, warning=FALSE}
knitr::opts_chunk$set(comment="#", fig.retina=2, crayon.enabled = TRUE)
set.seed(15)
# library(stats)
library(kableExtra)
library(knitr)
# library(formattable)
library(ggplot2)
library(ggthemes)
library(ggExtra)
# library(dplyr,warn.conflicts = FALSE)
# library(animation)
# library(gifski)
# library(latex2exp)
xaringanExtra::use_fit_screen()
```

## Learning Goals for Probability and Probability Distribution

- Demonstrate understanding of random variables

- Demonstrate understanding of characteristics of binomial distributions.

- Calculate accurate probabilities of discrete random variables and interpret them in a variety of settings.

---

## Random Variables

- A **random variable**, usually written $X$, is a variable whose values are numerical quantities of possible outcomes a random experiment.

- A **discrete random variable** takes on only a finite or countable number of distinct values.

  **Example:**

  - Rolling a fair dice, the number of dots on the top faces is a discrete random variables takes on the possible values: 1, 2, 3, ,4, 5, 6.
  - Flipping a fair coin 10 times, the number of heads is a discrete random variable takes on the possible values: 1, 2, 3, ..., 10.

- A **continuous random variable** takes on values which form an interval of numbers.

  **Example:**

  - The height of an randomly select 10 year-old boy in US is normally between 129 cm and 157 cm. So the height is a continuous random variable.
  - The measure the voltage at an randomly electrical outlet normally is between 118 and 122. So the measure of voltage is a continuous random variable.

---

## Practice: Discrete or continuous

Classify each random variable as either discrete or continuous.

1. The number of boys in a randomly selected three-child family.

2. The temperature of a cup of coffee served at a restaurant.

3. The number of math majors in randomly selected group of 10 students.

4. The amount of rain recorded in a small town one day.

---

## Probability Distributions

- The **probability distribution** of a discrete random variable $X$ is defined by the probability $P(X=x)$ associated with each possible value $x$ of the variable $X$. The function $p_X(x)=P(X=x)$ is called the **probability mass function**.

- A probability distribution of a discrete random variable is usually  characterized by a table of all possible values $X$ together with probabilities $P(X)$, or a probability histogram, or a formula.

- A random variable $X$ (discrete and continuous) always has a **cumulative distribution function**: $F_X(x)=P(X\leq x)$ (= $\sum\limits_{x_i\leq x} P(x_i)$ if $X$ is discrete).

---

## Basic Properties of Probability Distributions

- Basic rules of probability:

  - $0\leq P(X=x)\leq 1$.
  
  - the sum of all the probabilities is 1, that is $P(X\leq x_{max})=1$.
  
  - In particular, $0\leq F_X(x)\leq 1$.

  - The cumulative distribution function $F_X(x)$ is non-decreasing.

- The probability distribution can be recovered from its cumulative distribution function. Indeed, for a *discrete* random variable $X$, we have
  $$P(X=x_i)=P(X\le x_i)-P(X\le x_{i-1}),$$
  where $P(X\le x_i)=\sum\limits_{k=1}^i P(X=x_k)$.

---

## Example: Probability Distribution of Flipping Two Fair Coins

Let $X$ be the number of heads that are observed when tossing two fair coins.

1. Construct the probability distribution for $X$.
2. Find $P(X\le 1)$ and $P(X\le 2)$.

--

**Solution:** The possible values of the number of heads are $0$, $1$ and $2$. The probability distribution can be characterized by the following table:
.center[
| $x$   | 0   | 1   | 2   |
|-----|-----|-----|-----|
| $P(X=x)$ |0.25 | 0.5 | 0.25|
]

From the table, we may find the following cumulative distributions:
$$P(X\leq 1)=P(X=0)+P(X=1)=0.25+0.5=0.75.$$
$$P(X\leq 2)= P(X=0)+P(X=1)+P(X=2)=0.25+0.5+0.25=1.$$

---

## Example: Probability Histogram of an Unfair Coin

The probability distribution of an unfair coin is characterized by the following histogram. Find the probability of getting at most 1 head.
.center[
```{r echo=FALSE, fig.height = 3.5, fig.width = 5}
set.seed(10)
unfaircoin <- data.frame(Heads=rbinom(1000, 2, prob=0.4))

freqtb <- as.data.frame(table(unfaircoin$Heads))
prob<-round(freqtb$Freq/sum(freqtb$Freq),2)

ggplot(unfaircoin, aes(x=Heads), dpi=600) +
  geom_histogram(aes(y=stat(count) / sum(count)), binwidth=1, fill="lightblue", col="white",size=1)+
  labs(y = "Probability")+
  stat_bin(
    aes(y=stat(count) / sum(count),
    label=round(stat(count) / sum(count),2)),
    binwidth=1,
    geom="text",
    vjust=1.2
    )+
theme_bw(base_size = 12) +
theme(panel.grid=element_blank())
```
]

--

**Solution:** Let $X$ be the number of heads. From the probability histogram, we know that $P(X=0)=`r prob[1]`$, and $P(X=1)=`r prob[2]`$.

Then the probability of getting at most 1 head is
$$
P(X\le 1)=P(X=0)+P(X=1)=`r prob[1]`+`r prob[2]`=`r prob[1]+prob[2]`.
$$

---

## Practice: Probability Distribution of Rolling a Pair of Fair Dice

A pair of fair 6-sided dice were rolled. Let $X$ denote the sum of the number of dots on the top faces.

1. Construct the probability distribution of $X$.
2. Find the probability that $X$ takes an odd value.

---

## Practice: Probability Distribution of Bus Waiting Time

`r knitr::include_url("https://www.myopenmath.com/embedq2.php?id=916595&seed=2020&showansafter", height="560px")`

---

## Mean and Standard Deviation of a Discrete Random Variable

Let $X$ be a discrete random variable and $p_X(x)=P(X=x)$ the probability mass function.

- The **expected value** $E(X)$ (also called **mean** and denoted by $\mu$) of the discrete random variable $X$ is the number $$\mu=E(X)=\sum xp_X(x).$$

- The **variance** $\mathrm{Var}(X)$ (also denoted by $\sigma^2$) of the discrete random variable $X$ is the number $$\sigma^2=\mathrm{Var(X)}=\sum (x-E(X))^2p_X(x).$$

- The **standard deviation** $\sigma$ of a discrete random variable $X$ is the square root of its variance: $$\sigma=\sqrt{\sum (x-E(X))^2p_X(x)}.$$

<!-- --- -->

<!-- ## Some Properties of Expected Value (Optional) -->

<!-- - The expected value of a linear combination two random variables $X$ and $Y$ is the linear combination of their expected values, that is $$E(aX+bY)=aE(X)+bE(Y).$$ -->

<!-- - The expected value in general is not multiplicative, that is $E(XY)\ne E(X)E(Y)$. -->

<!-- - If the two random variable $X$ and $Y$ are independent, then $$E(XY)=E(X)E(Y).$$ -->

<!-- - The variance can be computed using expected values: -->
<!--   $$\mathrm{Var}(X)=E(X^{2})-(E[X])^{2}.$$ -->

<!-- - Let $X$ and $Y$ be two *independent* variables. Then the variance of a linear combination $aX+bY$ equals -->
<!-- $$\mathrm{Var}(aX+bY)=a^2\mathrm{Var}(X)+b^2\mathrm{Var}(Y)$$ -->

---

## Example: Expected Gain

One thousand raffle tickets are sold for $2 each. Each has an equal chance of winning. First prize is $500, second prize is $300, and third prize is $100. Find the expected value of net gain, and interpret its meaning.

**Solution:** Let $X$ denote the net gain from purchasing one ticket. The probability distribution for $X$ is  
.center[
| $x$    | 498    | 298    | 98     | -2       |
|------|--------|--------|--------|----------|
| $P(X=x)$ | $\frac{1}{1000}$ | $\frac{1}{1000}$ | $\frac{1}{1000}$ | $\frac{997}{1000}$ |
]

The expected gain is $$E(X)= 498\cdot \frac{1}{1000}+ 298\cdot\frac{1}{1000}+98\cdot \frac{1}{1000}+(-2)\cdot \frac{997}{1000}=-1.1,$$ which means that when buying one ticket, the buyer may expect a loss of $1.1.  

---

## Example: Waiting Time

The wait times (rounded to multiples of 5) in the cafeteria at a Community College has the following probability distribution. Find the expected waiting time and the standard deviation.
.center[
| $x$ (minutes) | 5        | 10       | 15      | 20       | 25       |
| -------------------- | -------- | -------- | -------- | -------- | -------- |
| $P(X=x)$             | 0.13 | 0.25 | 0.31 | 0.21 | 0.1 |
]

**Solution:** The expected waiting time is
$$E(X)= 5\cdot 0.13 +10\cdot 0.25+15\cdot 0.31+20\cdot 0.21 + 25\cdot 0.1 = 14.5.$$
The standard deviation is then
$$\scriptsize\begin{aligned}\sigma=&\sqrt{(5-14.5)^2\cdot 0.13 +(10-14.5)^2\cdot 0.25+(15-14.5)^2\cdot 0.31+(20-14.5)^2\cdot 0.21 + (25-14.5)^2\cdot 0.1}\\
\approx& 5.9.
\end{aligned}$$

---

## Example: Unfair Die (1/2)

The probability distribution of an unfair die is given in the following table.

| $x$ | 1| 2| 3| 4| 5| 6|
|---|---|---|---|---|---|
| $P(X=x)$ | 0.18| 0.12| $\,?\,$ |0.14|0.23|0.17|

1. Find $P(X=3)$.
2. Find the mean, variance and standard deviation of this probability distribution.

--

**Solution:** Since the sum of probabilities must be 1, we know that
$$
P(X=3)=1-(0.18+0.12+0.14+0.23+0.17)=0.16
$$

---

## Example: Unfair Die (2/2)

**Solution: (Continued)** The mean is the weighted sum
$$
% \scriptstyle
\begin{aligned}
\mu=&0.18\cdot 1+0.12\cdot 2+0.16\cdot 3+0.14\cdot 4+0.23\cdot 5+0.17\cdot 6\\
=&3.63.
\end{aligned}
$$

The variance is
$$
\scriptstyle
\begin{aligned}
\sigma^2=&0.18(1-3.63)^2+0.12(2-3.63)^2+0.16(3-3.63)^2+0.14(4-3.63)^2+0.23(5-3.63)^2+0.17(6-3.63)^2\\
=&3.0331.
\end{aligned}
$$

Therefore, the standard deviation is
$$
\sigma=\sqrt{\sigma^2}=\sqrt{3.0331}\approx 1.7416.
$$

---

## Practice: Lottery Tickets

Seven thousand lottery tickets are sold for $5 each. One ticket will win $2,000, two tickets will win $750 each, and five tickets will win $100 each. Let $X$ denote the net gain from the purchase of a randomly selected ticket.

1. Construct the probability distribution of $X$.
2. Compute the expected value $E(X)$ of $X$. Interpret its meaning.
3. Compute the standard deviation $\sigma$ of $X$.

.footmark[
Source: [https://saylordotorg.github.io/text_introductory-statistics/s08-02-probability-distributions-for-.html](https://saylordotorg.github.io/text_introductory-statistics/s08-02-probability-distributions-for-.html)
]

---

## Binomial Distribution

- A **binomial experiment** is a probability experiment satisfying:

  1. The experiment has a fixed number $n$ of independent trials.
  2. Each trial has only two possible outcomes: a success (S) or a failure (F).
  3. The probability $p$ of a success is the same for each trial.

- The discrete random variable $X$ counting the number of successes in the $n$ trials is the **binomial random variable**. We say $X$ has a **binomial distribution** with parameters $n$ and $p$ and write it as $X\sim B(n, p)$.

- For $X\sim B(n, p)$, the **probability of getting exactly $x$ successes in $n$ trials** is $$P(X=x)=B(x,n,p)={_n C_x} p^x(1-p)^{n-x}=\frac{n!}{(n-x)!x!}p^x(1-p)^{n-x},$$
where $n!=n(n-1)\cdots 1$, read as $n$ factorial, for $n>0$ and $0!=1.$

- The notation ${_n C_x}=\frac{n!}{(n-x)!x!}$ is read as $n$ choose $x$, which is the number of ways to choose $x$ objects from a set of $n$ objects.

---

## Example: Probability from Drawing a Card Multiple Times (1/2)

A card is randomly selected from a standard deck and replaced. This experiment is repeated a total of $5$ times.

- Find the probability of getting exactly $3$ clubs.
- Find the probability of getting at least $3$ clubs.

--

**Solution:** This is a binomial experiment. The number to total trials is $n=5$. The number of successes is $3$. The chance of a success is $p=\frac{13}{52}=\frac14$. Apply the binomial probability formula, we have
$$P(X=3)=\frac{5!}{3!2!} \left(\frac{1}{4}\right)^3\left(\frac34\right)^2=10\cdot\frac{9}{4^5}\approx 0.088.$$
The probability $P(X=3)$ can also be found from the binomial distribution table or by using the Excel function `BINOM.DIST(3,5,1/4,FALSE)`.

To probability of getting at least $3$ club is
$$P(X\geq 3) =1-P(X\leq 2)=1-(P(0)+P(1)+P(2))\approx 1-0.8965=0.1035.$$

---

## Example: Probability from Drawing a Card Multiple Times (2/2)
  
**Solution:(continued)**
To calculate $P(X\leq 2)$, we may also use the binomial distribution table or the Excel function `BINOM.DIST()`.

**Method 1:** As $n=5$ and $p=0.25$, we use the following portion of the cumulative binomial distribution table.
.center[Binomial Probability Table *n=5*

| n    | x    | 0.1    | 0.15   | 0.2    | 0.25   | 0.3    | 0.35   | 0.4    |
| ---- | ---- | ------ | ------ | ------ | ------ | ------ | ------ | ------ |
| 5    | 0    | 0.5905 | 0.4437 | 0.3277 | .red[0.2373] | 0.1681 | 0.116  | 0.0778 |
| 5    | 1    | 0.3281 | 0.3915 | 0.4096 | .red[0.3955] | 0.3602 | 0.3124 | 0.2592 |
| 5    | 2    | 0.0729 | 0.1382 | 0.2048 | .red[0.2637] | 0.3087 | 0.3364 | 0.3456 |
| 5    | 3    | 0.0081 | 0.0244 | 0.0512 | 0.0879 | 0.1323 | 0.1811 | 0.2304 |
]

$P(X\le 2) \approx 0.2373+0.3955+0.2637= 0.8965.$

**Method 2:** In Excel, $P(X\le 2)$ `=BINOM.DIST(2,5,0.25,TRUE)` $\approx 0.8965$.

???
In a calculator, use `binompdf(n, p, x)` for $P(X=x)$ and `binomcdf(n, p, x)` for $P(X\leq x)$.

---

## Practice: Find Probability of from a Binomial Distribution

Let $X$ be a binomial random variable with parameters $n = 5$, $p=0.2$. Find the probabilities

1. $P(X=3),$
2. $P(X<3),$
3. $P(X>3).$

---

## Practice: Machine Defect Rate

`r knitr::include_url("https://www.myopenmath.com/embedq2.php?id=7211&seed=2020&showansafter", height="560px")`

---

## Mean and Standard Deviation of Binomial Distribution (4/5)

- The mean of a binomial distribution of $n$ trials is $$\mu =\sum xP(X=x)=\sum x\cdot \dfrac{n!}{(n-x)!x!}p^x(1-p)^{n-x} = np.$$

- The variance of a binomial distribution of $n$ trials is $$\sigma^2 =\sum (x-np)^2P(X=x)=\sum x^2P(X=x)-(np)^2=np(1-p).$$  

- The standard deviation of a binomial distribution of $n$ trials is $$\sigma=\sqrt{np(1-p)}.$$

- We consider an event $E$ **unusual** if the probability $P(E)\leq 5\%$.

---

## Example: Expected Value and SD of Cracked Eggs

The probability that an egg in a retail package is cracked or broken is 0.02.

1. Find the average number of cracked or broken eggs in a one dozen carton.
2. Find the standard deviation.
3. Is getting at least two broken eggs unusual?

--

**Solution:** Since there are 12 eggs and the chance of getting a cracked egg is 0.02, the average number of cracked is
$$\mu =np=12\cdot 0.02=0.24.$$

The standard deviation is
$$\sigma=\sqrt{12\cdot 0.02\cdot(1-0.02)}\approx 0.4850.$$

Recall the Empirical rule: 95% data are within 2 standard deviation away from the mean. Since $2>0.24+2\cdot 0.4850$, the chance of getting at least two cracked eggs is less than 5%, which is considered as unusual.

---

## Practice: Quality of Grapefruit

Adverse growing conditions have caused 5% of grapefruit grown in a certain region to be of inferior quality. Grapefruit are sold by the dozen.

1. Find the average number of inferior quality grapefruit per box of a dozen.

2. A box that contains two or more grapefruit of inferior quality will cause a strong adverse customer reaction. Find the probability that a box of one dozen grapefruit will contain two or more grapefruit of inferior quality.

---

## Practice: Mean and SD of a Binomial Distribution

`r knitr::include_url("https://www.myopenmath.com/embedq2.php?id=916606&seed=2020&showansafter", height="560px")`

<!-- --- -->

<!-- # Extra Practice Problems -->

<!-- --- -->

<!-- ## Practice: Find the mean of a discrete probability distribution -->

<!-- `r knitr::include_url("https://www.myopenmath.com/embedq2.php?id=7197&seed=2020&showansafter", height="560px")` -->

<!-- --- -->

<!-- ## Practice: Find the standard deviation of a discrete probability distribution -->

<!-- `r knitr::include_url("https://www.myopenmath.com/embedq2.php?id=7198&seed=2020&showansafter", height="560px")` -->

<!-- --- -->

<!-- ## Practice: The number of sales of new employees -->

<!-- .pull-left[ -->
<!-- A company tracks the number of sales new employees make each day during a 100-day probationary period. The results for one new employee are shown at the right. -->

<!--   1. Find the probability of each outcome. -->
<!--   2. Construct a probability distribution table. -->
<!--   3. Find the mean of the probability distribution. -->
<!--   4. Find the variance and standard deviation. -->
<!-- ]<br> -->
<!-- .pull-right[ -->

<!-- | Sales per day $x$ | Number of days $f$ | -->
<!-- |-------------------|--------------------| -->
<!-- | 0                 | 16                 | -->
<!-- | 1                 | 19                 | -->
<!-- | 2                 | 15                 | -->
<!-- | 3                 | 21                 | -->
<!-- | 4                 | 9                  | -->
<!-- | 5                 | 10                 | -->
<!-- | 6                 | 8                  | -->
<!-- | 7                 | 2                  | -->
<!-- ] -->

<!-- --- -->

<!-- ## Practice: Probability within one SD of a discrete probability distribution -->

<!-- `r knitr::include_url("https://www.myopenmath.com/embedq2.php?id=14762&seed=2020&showansafter", height="560px")` -->


<!-- --- -->

<!-- ## Practice: Probability from a poll -->

<!-- `r knitr::include_url("https://www.myopenmath.com/embedq2.php?id=7210&seed=2020&showansafter", height="560px")` -->


<!-- --- -->

<!-- ## Practice: Chance of continuous successes of a type of surgery -->

<!-- A type of surgery has a 90% chance of success. The surgery is performed on three patients. Find the probability of the surgery being successful on exactly two patients. -->

---
class: center middle

# Lab Instructions in Excel

---

## Excel Functions for Binomial

Let $X$ be a binomial random variable with parameters $n$ and $p$, that is $X\sim B(n, p)$. In Excel, $P(X=x)$ is given by `BINOM.DIST(x, n, p, FALSE)` and $P(X\le x)$ is given by `BINOM.DIST(x, n, p, TRUE)`. You may click input function $f_x$ and then search `binom` to find the function.
